---
title: "L09 Themes"
subtitle: "Data Visualization (STAT 302)" 
author: "Bud Green"
pagetitle: "L09 Bud Green"
date: today

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-fold: show
    link-external-newwindow: true
    fig-dpi: 300

execute:
  warning: false

from: markdown+emoji 
---

## Overview

The goal of this lab is to play around with the theme options in `ggplot2`.


## Datasets

We'll be using the `cdc.txt` and `NU_admission_data.csv` datasets.

```{r warning = FALSE, message = FALSE}
#| label: load-pkgs-data

# load package(s)
library(tidyverse)
library(ggthemes)
library(patchwork)
library(showtext)

# read in the cdc dataset
cdc <- read_delim(file = "data/cdc.txt", delim = "|") |>
  mutate(
    genhlth = factor(
      genhlth,
      levels = c("excellent", "very good", "good", "fair", "poor"),
      labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")
      )
    )

# read in NU admission data
admin_data <- read_csv("data/NU_admission_data.csv") %>% 
  janitor::clean_names()
data(mpg)

# set seed
set.seed(2468)

# selecting a random subset of size 100
cdc_small <- cdc |> 
  slice_sample(n = 100)

showtext_auto()
```

## Exercise 1

Use the `cdc_small` dataset to explore several pre-set `ggthemes`. The code below constructs the familiar scatterplot of `weight` by `height` and stores it in `plot_01`. Display `plot_01` to observe the default theme. **Explore/apply, and display at least 7 other pre-set themes** from the `ggplot2` or `ggthemes` package. Don't worry about making adjustments to the figures under the new themes. Just get a sense of what the themes are doing to the original figure `plot_01`. 

There should be at least 8 plots for this task, `plot_01` is pictured below. Use `patchwork` or `cowplot` in combination with R yaml chunk options `fig-height` and `fig-width` (`out-width` and `fig-align` may be useful as well) to setup the 8 plots together in a user friendly arrangement. 

```{r}
#| label: ex-1-base-plot 
#| code-fold: true


# plot
plot_01 <- ggplot(
  data = cdc_small,
  aes(x = height, y = weight)
) +
  geom_point(size = 3, aes(shape = genhlth, color = genhlth)) +
  scale_y_continuous(
    name = "Weight in Pounds",
    limits = c(100, 275),
    breaks = seq(100, 275, 25),
    trans = "log10",
    labels = scales::label_number(
      accuracy = 1,
      suffix = " lbs"
    )
  ) +
  scale_x_continuous(
    name = "Height in Inches",
    limits = c(60, 80),
    breaks = seq(60, 80, 5),
    labels = scales::label_number(accuracy = 1, suffix = " in")
  ) +
  scale_shape_manual(
    name = "Health?",
    labels = c(
      "Excellent", "Very Good",
      "Good", "Fair", "Poor"
    ),
    values = c(17, 19, 15, 9, 4)
  ) +
  scale_color_brewer(
    name = "Health?",
    labels = c(
      "Excellent", "Very Good",
      "Good", "Fair", "Poor"
    ),
    palette = "Set1"
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification = c(1, 0)
  ) +
  labs(title = "CDC BRFSS: Weight by Height")
```

::: {.callout-tip icon=false}

## Solution

```{r}
#| label: ex-1-1
#| fig-show: hold
#| out-width: "50%"
#| layout-ncol: 2

plot_01
plot_01 + theme_excel()
plot_01 + theme_economist()
plot_01 + theme_wsj()
plot_01 + theme_calc()
plot_01 + theme_base()
plot_01 + theme_test()
plot_01 + theme_solarized()
```

:::

:::{.callout-important}
**Which theme or themes do you particularly like? Why?**
:::

::: {.callout-tip icon=false}

## Solution

These were all really ugly actually. the wsj one was the least disgusting looking one but still pretty bad.

:::

## Exercise 2

Using `plot_01` from Exercise 1 and the `theme()` function, attempt to construct the ugliest plot possible (example pictured below). Be creative! It should NOT look exactly like the example. Since the goal is to understand a variety of adjustments, you should use a **minimum** of 10 different manual adjustments within `theme()`.

::: {.callout-tip icon=false}

## Solution
```{r}
#| label: ex-2-ugly-plot

#base plot
plot_01 + theme_solarized_2() +
  theme(
    aspect.ratio = 2/9,
    plot.title = element_text(color = "orange", size = 9),
    axis.text = element_text(color = "cyan4", face = "italic"),
    strip.switch.pad.grid = unit(10, "lines"),
    panel.grid.major = element_line(color = "white", linewidth = 4, linetype = "dashed"),
    panel.spacing = unit(17, "inches"),
    legend.background = element_rect(fil = "pink"),
    legend.key.size = unit(0.00003, "inches"),
    axis.text.x = element_text(angle = 27, hjust = 0.7),
    axis.text.y = element_text(angle = 210),
    
  )

```

:::

## Exercise 3

We will be making use of your code from Exercise 3 on *L07 Layers*. Using the `NU_admission_data.csv` you created two separate plots derived from the single plot depicted in `undergraduate-admissions-statistics.pdf`. Style these plots so they follow a "Northwestern" theme. You are welcome to display the plots separately OR design a layout that displays both together (likely one stacked above the other). 

Check out the following webpages to help create your Northwestern theme:

- [Visual Identity](https://www.northwestern.edu/brand/visual-identity/index.html)
- [Color Palettes](https://www.northwestern.edu/brand/visual-identity/color-palettes/index.html)
- [Fonts & Typography](https://www.northwestern.edu/brand/visual-identity/fonts-typography/index.html) --- Need to use [substitute fonts](https://www.northwestern.edu/brand/visual-identity/fonts-typography/substitute-fonts/index.html)

**Additional requirement:**

Use a free non-standard font from google for the title. Pick one that looks similar to a Northwestern font.

:::{.callout-note}
I find this [blog post](https://cran.rstudio.com/web/packages/showtext/vignettes/introduction.html) to be extremely useful for adding fonts. Important packages for using non-standard fonts are `showtext`, `extrafont`, `extrafontdb`, and `sysfonts`. The last 3 generally just need to be installed (not loaded per session).
:::

::: {.callout-tip icon=false}

## Solution
```{r}
#| label: ex-3-NU-plot
#| fig-width: 9
#| fig-height: 4
#| fig-showtext: true


font_add_google("Quicksand")
#bar data
bar_data <- admin_data %>% 
  select(-contains("_rate")) %>% 
  pivot_longer(
    cols = -year,
    names_to = "category",
    values_to = "value"
  ) %>% 
  mutate(
    bar_labels = prettyNum(value, big.mark = ",")
  )

#building bar plot
bar_plot <- bar_data %>% 
  filter(year >= 2002) %>% 
  ggplot(aes(year, value)) +
  geom_col(
    mapping = aes(fill = category),
    width = 0.6,
    position = "identity"
  ) +
  geom_text(
    aes(label = bar_labels),
    size = 1.5,
    color = "white",
    vjust = 1,
    nudge_y = -200
  ) +
  scale_x_continuous(
    name = "Entering Year",
    breaks = 2002:2022,
    expand = c(0, 0.25)
  ) +
  scale_y_continuous(
    name = "Applications",
    limits = c(0, 70000),
    breaks = seq(0, 70000, 10000),
    labels = scales::label_comma(),
    expand = c(0,0)
  ) +
  scale_fill_manual(
    name = NULL,
    limits = c("applications", "admitted_students", "matriculants"),
    labels = c("Applications", "Admitted Students", "Matriculants"),
    values = c("#B6ACD1", "#836EAA", "#4E2A84")
  ) +
  theme_classic() +
  theme(
    legend.justification = c(0.5, 1),
    legend.position = "inside",
    legend.position.inside = c(0.5, 1),
    legend.direction = "horizontal",
    plot.title = element_text(hjust = 0.5, family = "Quicksand")
  ) +
  ggtitle("Northwestern University\nUndergraduate Admissions 2002-2022")
  
showtext_auto()
bar_plot
```

:::

## Challenge

:::{.callout-important icon=false}

Challenge is optional for all students, but we recommend trying them out!

:::

Using `cdc_small` dataset, re-create your own version inspired by the plot below.

**Must haves:**

-   Use two non-standard fonts (one for labeling the point and the other for the axes)
-   Use at least two colors (one for the added point, another for the rest of the points)
-   A curved arrow used to label the point

Using Bilbo Baggins' responses below to the CDC BRSFF questions, add Bilbo's data point to a scatterplot of `weight` by `height`.

-   `genhlth` - How would you rate your general health? **fair**
-   `exerany` - Have you exercised in the past month? **1=yes**
-   `hlthplan` - Do you have some form of health coverage? **0=no**
-   `smoke100` - Have you smoked at least 100 cigarettes in your life time? **1=yes**
-   `height` - height in inches: **46**
-   `weight` - weight in pounds: **120**
-   `wtdesire` - weight desired in pounds: **120**
-   `age` - in years: **45**
-   `gender` - m for males and f for females: **m**


:::{.callout-note}
Adding non-standard fonts can be an ***adventure***. I find this [blog post](https://cran.rstudio.com/web/packages/showtext/vignettes/introduction.html) to be extremely useful for adding fonts. Important packages for using non-standard fonts are `showtext`, `extrafont`, `extrafontdb`, and `sysfonts`. The last 3 generally just need to be installed (not loaded per session).
:::

*Hint:* 

- Create a new dataset (maybe call it `bilbo` or `bilbo_baggins`) using either `data.frame()` (base R - example in book) or `tibble()` (tidyverse - see help documentation for the function). Make sure to use variable names that exactly match `cdc`'s variable names. We have provided the `tidyverse` approach.
- Search google fonts to find some free fonts to use (can get free fonts from other locations)

::: {.callout-tip icon=false}

## Solution

```{r}
#| label: challenge

# build dataset for Bilbo Baggins
bilbo <- tibble(
  genhlth  = "fair",
  exerany  = 1,
  hlthplan = 0,
  smoke100 = 1,
  height   = 46,
  weight   = 120,
  wtdesire = 120,
  age      = 45,
  gender   = "m"
)

#fonts
font_add_google("Boogaloo")
font_add_google("Lobster")
#cdc_small plot
cdc_small_plot <- cdc_small %>% 
  ggplot(aes(x = height, y = weight)) +
  geom_point(color = "#ffaa00") +
  scale_x_continuous(
    name = "Height (in)",
    limits = c(45, 80)
  ) +
  scale_y_continuous(
    name = "Weight (lbs)",
    limits = c(100, 400)
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(family = "Boogaloo", size = 20)
  ) +
  geom_curve(
    aes(x = 65, y = 370, xend = 46, yend = 123),
    curvature = 0.4,
    arrow = arrow(length = unit(0.2, "inches")),
    color = "#0065FE"
  ) +
  geom_point(data = bilbo, aes(x = height, y = weight), color = "#0065FE") +
  geom_text(
    aes(label = "Bilbo Baggins", family = "Lobster"),
    color = "#0065FE", 
    x = 65,
    y = 375,
    hjust = 0,
    vjust = 1,
    size = 20
  )
 
cdc_small_plot
```

:::
