---
title: "L03 ggplot 2"
subtitle: "Data Visualization (STAT 302)" 
author: "Bud Green"
pagetitle: "L03 Bud Green"
date: today

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    link-external-newwindow: true
    fig-dpi: 150

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---


## Overview

The goal of this lab is to continue the process of unlocking the power of `ggplot2` through constructing and experimenting with a few basic plots.

## Datasets

We will be using the BRFSS survey which was introduced in the last lab. The data was supplied in `cdc.txt` file and should be in your `/data` subdirectory. If not, you need to download the last lab to get this data file. As a reminder, the dataset contains 20,000 complete observations/records of 9 variables/fields, described below.

* `genhlth` - How would you rate your general health? (excellent, very good, good, fair, poor)
* `exerany` - Have you exercised in the past month? (`1 = yes`, `0 = no`) 
* `hlthplan` - Do you have some form of health coverage? (`1 = yes`, `0 = no`) 
* `smoke100` - Have you smoked at least 100 cigarettes in your life time? (`1 = yes`, `0 = no`) 
* `height` -  height in inches
* `weight` - weight in pounds
* `wtdesire` - weight desired in pounds
* `age` - in years
* `gender` - `m` for males and `f` for females

```{r warning=FALSE, message=FALSE, include=FALSE}
#| label: load-packages

# Load package(s)
library(tidyverse)

# Load data
cdc = read_delim(file = "data/cdc.txt", delim = "|") %>% 
  mutate(
    genhlth = factor(
      genhlth,
      levels = c("excellent", "very good", "good", "fair", "poor"))
  )
```

## Exercise 1

Using the `cdc` dataset, we want to look at the relationship between height and weight. Recreate the following graphics as precisely as possible.

### Plot 1

*Hints:*

- Transparency is 0.2
- Minimal theme

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex-1-1


ggplot(cdc, aes(height, weight, color = gender)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE) +
  labs(x = "Height (in)", y = "Weight (lbs)") +
  theme_minimal()
```

:::

### Plot 2

*Hints:*

- linewidth = 0.7

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex-1-2

ggplot(cdc, aes(height, weight, color = gender)) +
  geom_density_2d(linewidth = 0.7) +
  labs(x = "Height (in)", y = "Weight (lbs)") +
  theme_minimal()
```

:::

### Plot 3

*Hints:*

- `bins` set to 35

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex-1-3

ggplot(cdc, aes(height, weight)) +
  geom_hex(bins = 35) +
  labs(x = "Height (in)", y = "Weight (lbs)") +
  theme_minimal()
```
:::

### Plot 4

*Hints:*

- use a stat layer, not a geom layer
- `geom = "polygon"` 

::: {.callout-tip icon="false"}
## Solution
```{r}
#|label: ex-1-4

ggplot(cdc, aes(height, weight)) +
  stat_density_2d(
    geom = "polygon",
    mapping = aes(fill = after_stat(level)),
    show.legend = FALSE
    ) +
  facet_wrap(~ gender) +
  labs(x = "Height (in)", y = "Weight (lbs)") +
  theme_minimal()
```

:::

## Exercise 2

Using the `cdc_means` dataset derived from the `cdc` dataset, recreate the following graphic as precisely as possible.

*Hints:*

- Hex color code `#56B4E9`
- 95% confidence intervals (1.96 or `qnorm(0.975)`)
- Some useful values: 0.1, 0.7


```{r} 
#| label: ex-2-plot

# data wrangling
# calc mean and se for CI
cdc_means = cdc %>% 
  mutate(wtloss = weight - wtdesire) %>% 
  group_by(genhlth) %>%
  summarize(
    mean = mean(wtloss),
    se = sd(wtloss) / sqrt(n())
  ) %>% 
  mutate(genhlth = fct_reorder(factor(genhlth), desc(mean)))
```


::: {.callout-tip icon="false"}
## Solution
```{r}
#| label: ex-2-1

ggplot(data = cdc_means, aes(genhlth, mean)) +
  geom_bar(stat = "identity", fill = "#56B4E9", width = 0.7) +
  geom_errorbar(aes(ymin = mean - (qnorm(0.975) * se), ymax = mean + (qnorm(0.975) * se)), width = 0.1) +
  theme_minimal() +
  labs(x = "General Health", y = "Mean desired weight loss in lbs")
```

:::

## Exercise 3

Using the `cdc_weight_95ci` dataset derived from the `cdc` dataset, recreate the following graphic as precisely as possible.

*Hints:*

- Useful values: 0.1, 0.5
- Need to know CI formula

```{r}
#| label: ex-3-plot

# data wrangling
# calculate mean, se, and margin of error for CI formula
cdc_weight_95ci <- cdc %>% 
  group_by(genhlth, gender) %>% 
  summarise(
    mean_wt = mean(weight),
    se = sd(weight) / sqrt(n()),
    moe = qt(0.975, n() - 1) * se
  ) %>% 
  ungroup()
```


::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex-3-1

ggplot(data = cdc_weight_95ci, aes(mean_wt, genhlth)) +
  geom_point(aes(color = genhlth), position = position_dodge(width = 5)) +
  geom_errorbar(aes(xmin = mean_wt - moe, xmax = mean_wt + moe, color = genhlth), width = 0.1) +
  
  facet_grid(factor(gender, levels = c("m", "f")) ~ ., scales = "free_y", switch = "y") +
  labs(x = "Weight (lbs)", y = "Gender", color = paste("General health", "\n(self reported)")) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        panel.grid.major.y = element_line(size = 0.5)
        )
```
:::
